---
title: "HW 21" 
author: "Harsha"
date: "2024-04-01"
output: 
   html_document:
     code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#global rules for entire markdown
```


```{r fig.width=4, fig.height=3}
library(dplyr)
library(ggplot2)
library(tukeyedar)

options(scipen = 99)
#load data
dat <- read.csv("census_income.csv")

#extract income
income  <- dat %>% 
  arrange(income) %>% 
  pull(income)


# Comparing income to a (unit) normal qq plot
ggplot(dat, aes(sample=income)) + stat_qq(distribution=qnorm) +
  stat_qq_line( col = "blue") +
  xlab("Unit normal quantile") + ylab("Income") + 
  ggtitle("Actual income vs Unit normal quantile") 


```

When comparing each state's income to a normal distribution, we focus where the points follow the fitted line at the core of the dataset (i.e. between -1 and 1 standard deviations). In our case, our income data is approximately normal if most of our income points between -1 and 1 standard deviations fall on the blue line of what the income would be if they followed a normal distribution. We can see that between 50K and 70K, which is above and below a standard deviation from the mean of 0 in a standard normal distribution, most of the income points are relatively close to the line of what the points would be in a normal distribution. It should be noted that most of these income points seem to be a little less than what they would be in a normal population. We can examine further if this is a concern by simulating random normal income values and comparing that to our actual income values.




```{r fig.width=7, fig.height=7}
# Simulate values from a normal distribution
sim <- data.frame(sample = paste0("Sample",1:16),
                  value  = round(rnorm(length(income)*16, 
                                       mean = mean(income), sd = sd(income))))

# Generate q-q plots of the simulated values
ggplot(sim, aes(sample = value)) + stat_qq(distribution = qnorm) +
  stat_qq_line(line.p = c(0.25, 0.75), col = "blue") +
  xlab("Unit normal quantile") + ylab("Simulated normals") +
  facet_wrap(~ sample, nrow = 4) 
```

The state income normal q-q plots does not look much different from some of these 16 simulated plots. Albeit the simulated plots above generally have more income points close the blue line, I believe the prior graph of our actual income values against the unit normal quantile has enough points close enough to the blue line in this graph. The above plots look very similar to that generated before. In fact, the prior graph seems to be more normal than the graphs of sample 11 and 12, which has considerable points deviating from the line. The state income distributions can be characterized by a normal distribution.
