---
title: "HW 19" 
author: "Harsha"
date: "2024-03-21"
output: 
   html_document:
     code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#global rules for entire markdown
```

#### Part 1

```{r fig.width=5, fig.height=3.5}
# Load packages
library(dplyr)
library(tukeyedar)
library(ggplot2)

dat <- read.csv("https://mgimond.github.io/ES218/Data/Shanghai.csv")

#Categorize by century 
dat1 <- dat %>% 
  mutate(Century = if_else(Year < 2000, "1800s", "2000s"))

#Create boxplots of temperature by century 
ggplot(dat1, aes(x = Century, y = Temp)) + geom_boxplot(fill = "bisque") +
  xlab("Century") + ylab("Temperature") + 
  ggtitle("Boxplot of Temperature by Century in Shanghai")

```


This is a dataset of temperatures in Shanghai in the 1800s and 2000s. First, I took this long table format data and added a column that will categorize each datapoints according to the century it belongs. The lower 25% of the temperatures (up to Q1) is generally lower for the 19th century than the 21st. The minimum for the 19th century, which is almost 0 degrees, is lower than that of the 19th century of around 0.7 degrees. The median or 50th percentile in the 21st century of around 18 degrees is higher than that in the 19th century of around 16.5 degrees. The top 25% of the data from the 75th percentile on also differs. For the 21st century the 75th percentile is around 25 degrees, while for the other it is around 24 degrees. The middle 50% of the temperatures or inner quartiles tends to be slightly higher in the 2000s than 1800s. 

#### Part 2

```{r fig.height=4, fig.width=4, results='hide'}
# A Tukey mean-difference plot for the centuries
eda_qq(dat1, Temp, Century) #Before adding fucntion
eda_qq(dat1, Temp, Century, fx = "x * 0.9781 + (1.3785)")
```

Based on the suggested f(x) from the eda_qq function and the QQ plot, there seems to be an multiplicative and additive relationship between both centuries. On first inspection, the 2000s seem to have higher temperatures than the 1800s at a constant rate. The data points are all shifted consistently above the line of temperatures in 1800s equals that in the 2000s. However, looking closer it is evident that this difference is highest when the temperatures is around 15 or 2 degrees (average, cold days respectaviely) in the 1800s. In fact, at warmer temperatures of more than 30 degrees (very hot days), the trend switches, where the 1800s actually had slightly higher temperatures. There seem to be different patterns for cooler temperatures than for warmer temperatures. Since the difference in temperatures is not constant, there may be an multiplicative relationship as well. Looking at the Tukey M-D plot, we can see that indeed the relationship is a combination of both since now the points fall on a line with the equation Temperature in 2000s = temperature in 1800s * 0.9781 + (1.3785). 
